---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Loading and preprocessing the data
```{r}
library(tidyverse)
unzip("activity.zip")
active <- read.csv("activity.csv")
active$date <- as.Date(active$date, format = "%Y-%m-%d")
```
## What is mean total number of steps taken per day?
```{r}
steps <- active %>%
        group_by(date) %>%
        summarize(mean_steps = mean(steps, na.rm = TRUE), 
                  total_steps = sum(steps, na.rm = TRUE),
                  median_steps = median(steps, na.rm = TRUE))
mean_before_imputatiaon <- mean(steps$total_steps, na.rm = TRUE)
median_before_imputation <- median(steps$total_steps, na.rm = TRUE)
mean_before_imputatiaon
median_before_imputation
```
Before imputation to NAs, the mean number of steps per day is 9354.23 and the median is 10395.
```{r Figure 1: Histogram of Total Daily Steps Before Imputing NAs}
## Make a histogram of the total number of steps taken each day
steps_hist <- ggplot(steps, aes(total_steps))
steps_hist + geom_histogram(na.rm = TRUE) + 
        labs(title = "Histogram of Total Steps Per Day",
             subtitle = "https://github.com/rdpeng/RepData_PeerAssessment1",
             x = "Total Steps Per Day", 
             y = "Days") +
        scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10), labels = c(0, 2, 4, 6, 8, 10))
```

## What is the average daily activity pattern?
```{r Figure 2: Time Series Plot of Average Daily Activity by Five Minute Intervals}
intervals <- active %>%
        group_by(interval) %>%
        summarize(mean_steps = mean(steps, na.rm = TRUE))
plot_intervals <- ggplot(intervals, aes(interval, mean_steps))
plot_intervals + geom_line() + 
        labs(title = "Average Steps Per Five-Minute Interval Per Day",
             subtitle = "https://github.com/rdpeng/RepData_PeerAssessment1",
             x = "Minutes",
             y = "Steps Per Five Minutes")
```
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps? 
```{r}
max_steps <- max(intervals$mean_steps, na.rm = TRUE)
max_step_interval <- intervals$interval[intervals$mean_steps == max_steps]
hour_of_day <- max_step_interval %/% 60 
minute_of_hour <- max_step_interval %% 60
hour_of_day
minute_of_hour
```
The five minute interval with maximum steps is from 835 to 840 minutes (13h55 to 14h).
## Imputing missing values
```{r}
### (i.e. the total number of rows with NAs)
sum(!complete.cases(active))
sum(is.na(active$steps))
```
There are 2304 cases with an NA and all of them are in the steps variable. 
```{r Imputation of NAs as the mean for the five-minute period.}
active_not_na <- active %>%
        select(steps, date, interval) %>%
        group_by(interval) %>%
        mutate(mean_steps = mean(steps, na.rm = TRUE))
active_not_na$steps[is.na(active_not_na$steps)] <- 
        active_not_na$mean_steps[is.na(active_not_na$steps)]
```
Histogram of Total Daily Steps After Imputing NAs
```{r Figure 3: Histogram of Total Daily Steps After Imputing NAs}
steps_not_na <- active_not_na %>%
        group_by(date) %>%
        summarize(total_steps = sum(steps))
steps_not_na_hist <- ggplot(steps_not_na, aes(total_steps))
steps_not_na_hist + geom_histogram(na.rm = TRUE) + 
        labs(title = "Histogram of Total Steps Per Day, NA Imputed",
             subtitle = "https://github.com/rdpeng/RepData_PeerAssessment1",
             x = "Total Steps Per Day", 
             y = "Days") +
        scale_y_continuous(breaks = c(0, 2, 4, 6, 8, 10), labels = c(0, 2, 4, 6, 8, 10))
```
## Calculate and report the mean and median total number of steps taken per day. 
```{r}
mean_after_imputation <- mean(steps_not_na$total_steps)
median_after_imputation <- median(steps_not_na$total_steps)
mean_after_imputation
median_after_imputation
mean_after_imputation - mean_before_imputatiaon
median_after_imputation - median_before_imputation
```
After imputation, the mean and median are the same (10766.19). The decimal in a median results from imputation of means to NAs. The mean is 1411.959 larger and the median is 371.1887 larger than before imputation. 

## Are there differences in activity patterns between weekdays and weekends?
```{r Differences in weekday and weekend activity}
## Create a new factor variable in the dataset with two levels – “weekday” and “weekend”
weekday_index <- weekdays(active_not_na$date)
active_not_na$weekday[weekday_index %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")] <- "weekday"
active_not_na$weekday[weekday_index %in% c("Saturday", "Sunday")] <- "weekend"
active_not_na$weekday <- as.factor(active_not_na$weekday)
```
Panel Time Series of Weekday and Weekend Activity
```{r Figure 4: Time Series Panel Plot of Weekday and Weekend Activity}
weekday_steps <- active_not_na %>%
        group_by(weekday, interval) %>%
        summarize(average_steps = mean(steps))
weekday_plot <- ggplot(weekday_steps, aes(interval, average_steps))
weekday_plot + geom_line() + 
        labs(title = "Comparison of Weekday and Weekend Steps Per Five-Minute Interval Per Day",
             subtitle = "https://github.com/rdpeng/RepData_PeerAssessment1",
             x = "Minutes",
             y = "Steps Per Five Minutes") +
        facet_grid(weekday~.)
